Pre working group data exploration: Mainly wrt coral cover

### 2024-07-29 julianna_data_exploration.Rmd
Calculated the change in algal cover over the next year based on the richness/diversity in that year

@@

>>>>

- Would need to start in 2008 because no change to 2007

## Wrangle

Might just brute force this...

```{r get the change in algal cover each year}
megaQuadDF %>% 
  select(year, site, habitat, site_habitat, transect, quadrat, island_side, AlgSum) %>% 
  pivot_wider(names_from = year,
              values_from = AlgSum) %>% 
  # calculate changes FROM the previous year 
  # this means that a change FROM 2007 to 2008 will be joined with diversity in 2007 
  # that means that a positive correlation between richness in 2007 and "ChFrom2007" (etc) == more diverse communities are more likely to INCREASE in cover/less likely to decrease in cover (not necessarily stability)
  mutate(AlgChFrom2007 = `2008` - `2007`,
         AlgChFrom2008 = `2009` - `2008`,
         AlgChFrom2009 = `2010` - `2009`,
         AlgChFrom2010 = `2011` - `2010`,
         AlgChFrom2011 = `2012` - `2011`,
         AlgChFrom2012 = `2013` - `2012`,
         AlgChFrom2013 = `2014` - `2013`,
         AlgChFrom2014 = `2015` - `2014`,
         AlgChFrom2015 = `2016` - `2015`,
         AlgChFrom2016 = `2017` - `2016`,
         AlgChFrom2017 = `2018` - `2017`,
         AlgChFrom2018 = `2019` - `2018`,
         #AlgChFrom2007 = `2021` - `2019`, # maybe skip because this isn't comparable?
         AlgChFrom2021 = `2022` - `2021`, 
         AlgChFrom2022 = `2023` - `2022`
         ) -> algalCh
```

There are **NA's in 2021**--confirm those are just missing quadrats.

Now get it back into long/joinable format (probably a more elegant way to do this):

```{r get into long format and join}
startChA <- grep("AlgChFrom2007", colnames(algalCh))
endtChA <- grep("AlgChFrom2022", colnames(algalCh))

algalCh %>% 
  # get the columns we need
  select(c(site, habitat, site_habitat, transect, quadrat, island_side, colnames(algalCh)[startChA:endtChA])) %>% 
  pivot_longer(cols = colnames(algalCh)[startChA:endtChA],
               names_to = "Change_year",
               values_to = "Change_in_percent"
               ) %>% 
  # match with correct year
  mutate(year = case_when(
    Change_year == "AlgChFrom2007" ~ 2007,
    Change_year == "AlgChFrom2008" ~ 2008,
    Change_year == "AlgChFrom2009" ~ 2009,
    Change_year == "AlgChFrom2010" ~ 2010,
    Change_year == "AlgChFrom2011" ~ 2011,
    Change_year == "AlgChFrom2012" ~ 2012,
    Change_year == "AlgChFrom2013" ~ 2013,
    Change_year == "AlgChFrom2014" ~ 2014,
    Change_year == "AlgChFrom2015" ~ 2015,
    Change_year == "AlgChFrom2016" ~ 2016,
    Change_year == "AlgChFrom2017" ~ 2017,
    Change_year == "AlgChFrom2018" ~ 2018,
    Change_year == "AlgChFrom2019" ~ 2019,
    Change_year == "AlgChFrom2021" ~ 2021,
    Change_year == "AlgChFrom2022" ~ 2022,
    TRUE ~ -9999
  )) %>%
  full_join(megaQuadDF) %>% 
  mutate(habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17")),
         island_side = factor(island_side, levels = c("north", "east", "west"))) -> algChYr

```{r plot richness}
algChYr %>% 
  ggplot(aes(x = Richness, y = Change_in_percent, group = Richness)) +
  geom_boxplot() +
  theme_classic() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_grid(island_side~habitat)
```

```{r plot abs(change) ~ richness}
algChYr %>% 
  ggplot(aes(x = Richness, y = abs(Change_in_percent), group = Richness)) +
  geom_boxplot() +
  theme_classic() +
  facet_grid(island_side~habitat)
```

```{r plot Shannon}
algChYr %>% 
  ggplot(aes(x = Shannon, y = Change_in_percent)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth() +
  theme_classic() +
  facet_grid(island_side~habitat)
```

```{r plot Shannon ~ abs(change)}
# pdf("change_in_cov_Shann.pdf", height = 7, width = 10)

algChYr %>% 
  ggplot(aes(x = Shannon, y = abs(Change_in_percent), color = year)) +
  geom_point() +
  geom_smooth(color = "coral") +
  theme_classic() + # kind of negative
  facet_grid(island_side~habitat)

@@

# Nic's plot of change through time

Want to plot how each spp changes in abundance through time just to get an idea

## Wrangling

```{r}
megaQuadDF
```
